---
title: "MSB 1015 Scientific Programming - Assignment 2"
author: "Ammar Ammar"
date: "October 12, 2019"
output:
  html_document:
    highlight: tango
    theme: flatly
---

## Dependencies

First, you have to check if the required packages are already installed. If not, install them:

```{r installPackages}
requiredPackages <- c("WikidataQueryServiceR", # run SPARQL query against Wikidata
                      "rJava",  # required for rcdk
                      "rcdk",   # cheminformatics library
                      "randomForest", # Random forest machine learning library
                      "caret",  # Classification And REgression Training 
                      "dplyr",  # data manipulation library
                      "stringr", # string manipulation library
                      "pls",   # Partial Least Squares regression library
                      "measurements",  # library to convert to many units
                      "DT") # generate interactive data tables

for (pkg in requiredPackages) { 
    if(! pkg %in% row.names(installed.packages())) install.packages(pkg)
}
```

Load the required packages:

```{r loadPackages, results = 'hide', message=FALSE, warning=FALSE}
library(WikidataQueryServiceR)
library(rJava)
library(rcdk)
library(randomForest)
library(caret)
library(dplyr)
library(stringr)
library(pls)
library(measurements)
library(DT)
```

## Load Data

Prepare a SPARQL query to get alkanes with their smiles and boiling points from Wikidata


```{r sparql}
sparql_query <- 'SELECT ?comp ?compLabel ?smile ?bp ?bpUnit ?bpUnitLabel WHERE {
  ?comp wdt:P31/wdt:P279* wd:Q41581 ;
        wdt:P233 ?smile ;
        p:P2102 [
          ps:P2102 ?bp ;
          psv:P2102/wikibase:quantityUnit  ?bpUnit
        ] .
  SERVICE wikibase:label { bd:serviceParam wikibase:language     "[AUTO_LANGUAGE],en". }
}'

results <- query_wikidata(sparql_query)

summary(results)

```

## Preprocess the data

Next, process the boiling point by looking at the unit label and convert from Fahrenheit and Celesius to Kelvin. Then, select only the alkan label, SMILE and boiling point into a new data frame:

```{r filterResults}

# if the unit in Fagrenheit or Celesius, use conv_unit function from the "measurements" library to convert the temprature to Kelvin
alkanesData <- results %>%
               mutate(bp = case_when(bpUnitLabel == "degree Fahrenheit" ~ conv_unit(bp ,"F","K"),
                                     bpUnitLabel == "degree Celsius" ~ conv_unit(bp ,"C","K"),
                                     TRUE ~ bp)) %>%
               select(compLabel, smile, bp)

# output the final alkanes data frame as interactive HTML table with sort, search and pagination functions
datatable(alkanesData[,c(1,3)], rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=T))
```

## Calcualte the features (The Descriptors)

Parse the SMILE strings of the compounds and calculate all the descriptors for each one of them. Then, bind the boiling points column to the descriptors data frame:

```{r calcDescriptors}
# parse the SMILE strings into IAtomCOntainer object for each Alkane compound
mols <- parse.smiles(alkanesData$smile)

# get a list of all descriptors available in CDK
descriptorsNames <- get.desc.names(type = "all")

# calculate the desctiptors for all the alkanes
descriptors <- eval.desc(mols, descriptorsNames)

# combine the descriptors values and the boiling points in one data frame
featuresDataframe <- cbind(descriptors, alkanesData$bp)

# change the name of the boiling point column into a more readable name
colnames(featuresDataframe)[ncol(featuresDataframe)] <- "bp"
```

## Clean the data

Remove the descriptors and molecules which have NA value(s):

```{r removeNAs}
featuresDataframe <- featuresDataframe[,colSums(is.na(featuresDataframe)) < nrow(featuresDataframe)]

featuresDataframe <- featuresDataframe[rowSums(is.na(featuresDataframe)) == 0,]

```

## First Model: Prtial Least Squares


### Split the data

Randomly sample 20% of the total number of alkanes then create two datasets (train set and test set):

```{r splitDataset}

# set a set to generate the same random numbers (for reproducibility)
set.seed(123)

# select ~20% test set
test_selection <- sample(nrow(featuresDataframe), 0.2 * nrow(featuresDataframe))

# create train and test datasets from the features dataframe
traint_set <- featuresDataframe[-test_selection,]
test_set <- featuresDataframe[test_selection,]
```

### Build the model

Here, we fit a partial least square model on the training set. This fits a model with 10 components, and includes leave-one-out (LOO) cross-validated predictions:

```{r fitPLS}
pls_model <- plsr(bp ~ ., ncomp = 10, data = traint_set, validation = "LOO")

summary(pls_model)
```

The validation results in the following plot are Root Mean Squared Error of Prediction (RMSEP). There are two cross-validation estimates: CV is the ordinary CV estimate, and adjCV is a bias-corrected CV estimate. (For a LOO CV, there is virtually no difference).
It is often simpler to judge the RMSEPs by plotting them:
```{r plotRMSEP, fig.align='center'}

#plot estimated RMSEPs as functions of the number of component.
#The legendpos argument adds a legend at the indicated position
plot(RMSEP(pls_model), legendpos = "topright", main="RMSEP vs. number of components")
```
```{r printRMSEP}
RMSEP(pls_model)
```


**Three components seem to be enough**. Once the number of components has been chosen, one can inspect different aspects of the fit by plotting predictions, scores, correlation and coeffecients .. etc.

Left plot visualizes the regression coefficients for the first three components and shows the impact of the predictors on the boiling point. The X axis is the predictor index and the Y axis is the regression coeffecient.

Right plot shows the cross-validated predictions with three components versus measured values. We have chosen to draw a target line using the paramter line=TRUE.


```{r plotModel, out.width=c('50%', '50%'), fig.show='hold'}

plot(pls_model, plottype = "coeff", ncomp = 3, main="Boiling Point PLS coeffecients plot")

plot(pls_model, plottype = "prediction", ncomp = 3, main="Boiling Point PLS CV predicion plot", line=TRUE, pch=16,col="blue")
```

### Apply prediction

The fitted model is used to predict the response values of the test dataset. The following predicts the responses for the 27 alkanes in test_set, using three components:

```{r prediction}
# use the fitted model to predict the values of the test dataset using 3 components
predicted_test <- predict(pls_model, ncomp = 3, newdata = test_set)
```

R^2 is a statistical measure that represents the proportion of the variance for a dependent variable that's explained by an independent variable or variables in a regression model. By plotting the R^2 values againt the number of components we can see that for three components, ~99% of the variance is explained by 3 components.

```{r predictionPlots, out.width=c('50%', '50%'), fig.show='hold'}
# plot the R2 values as a measure of prediction performance
plot(R2(pls_model, newdata = test_set), main="Boiling point PLS, R2 vs. number of components")

# plot the predictet value vs. the measured values of the test set
predplot(pls_model, ncomp = 3, newdata = test_set, line = TRUE, pch=16,col="blue", main="Boiling point, 3 components, test set")
```

Also, we report the RMSEP plot for the test set

```{r predictionRMSEP, fig.align='center'}

# plot the RMSEP for the test set
plot(RMSEP(pls_model, newdata = test_set), legendpos = "topright")

```

```{r printRMSEPtest}

RMSEP(pls_model, newdata = test_set)
```

**R2 values**

```{r printR2test}

R2(pls_model, newdata = test_set)

```