---
title: "MSB 1015 Scientific Programming - Assignment 2"
author: "Ammar Ammar"
date: "October 12, 2019"
output:
  html_document:
    highlight: tango
    theme: flatly
---

## Dependencies

First, you have to check if the required packages are already installed. If not, install them:

```{r installPackages}
requiredPackages <- c("WikidataQueryServiceR", # run SPARQL query against Wikidata
                      "rJava",  # required for rcdk
                      "rcdk",   # cheminformatics library
                      "randomForest", # Random forest machine learning library
                      "caret",  # Classification And REgression Training 
                      "dplyr",  # data manipulation library
                      "stringr", # string manipulation library
                      "pls",   # Partial Least Squares regression library
                      "measurements",  # library to convert to many units
                      "DT") # generate interactive data tables

for (pkg in requiredPackages) { 
    if(! pkg %in% row.names(installed.packages())) install.packages(pkg)
}
```

Load the required packages:

```{r loadPackages, results = 'hide', message=FALSE, warning=FALSE}
library(WikidataQueryServiceR)
library(rJava)
library(rcdk)
library(randomForest)
library(caret)
library(dplyr)
library(stringr)
library(pls)
library(measurements)
library(DT)
```

## Load Data

Prepare a SPARQL query to get alkanes with their smiles and boiling points from Wikidata


```{r sparql}
sparql_query <- 'SELECT ?comp ?compLabel ?smile ?bp ?bpUnit ?bpUnitLabel WHERE {
  ?comp wdt:P31/wdt:P279* wd:Q41581 ;
        wdt:P233 ?smile ;
        p:P2102 [
          ps:P2102 ?bp ;
          psv:P2102/wikibase:quantityUnit  ?bpUnit
        ] .
  SERVICE wikibase:label { bd:serviceParam wikibase:language     "[AUTO_LANGUAGE],en". }
}'

results <- query_wikidata(sparql_query)

summary(results)

```

## Preprocess the data

Next, process the boiling point by looking at the unit label and convert from Fahrenheit and Celesius to Kelvin. Then, select only the alkan label, SMILE and boiling point into a new data frame:

```{r filterResults}

# if the unit in Fagrenheit or Celesius, use conv_unit function from the "measurements" library to convert the temprature to Kelvin
alkanesData <- results %>%
               mutate(bp = case_when(bpUnitLabel == "degree Fahrenheit" ~ conv_unit(bp ,"F","K"),
                                     bpUnitLabel == "degree Celsius" ~ conv_unit(bp ,"C","K"),
                                     TRUE ~ bp)) %>%
               select(compLabel, smile, bp)

# output the final alkanes data frame as interactive HTML table with sort, search and pagination functions
datatable(alkanesData[,c(1,3)], rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=T))
```

## Calcualte the features (The Descriptors)

Parse the SMILE strings of the compounds and calculate all the descriptors for each one of them. Then, bind the boiling points column to the descriptors data frame:

```{r calcDescriptors}
# parse the SMILE strings into IAtomCOntainer object for each Alkane compound
mols <- parse.smiles(alkanesData$smile)

# get a list of all descriptors available in CDK
descriptorsNames <- get.desc.names(type = "all")

# calculate the desctiptors for all the alkanes
descriptors <- eval.desc(mols, descriptorsNames)

# combine the descriptors values and the boiling points in one data frame
featuresDataframe <- cbind(descriptors, alkanesData$bp)

# change the name of the boiling point column into a more readable name
colnames(featuresDataframe)[ncol(featuresDataframe)] <- "bp"
```

## Clean the data

Remove the descriptors and molecules which have NA value(s):

```{r removeNAs}
featuresDataframe <- featuresDataframe[,colSums(is.na(featuresDataframe)) < nrow(featuresDataframe)]

featuresDataframe <- featuresDataframe[rowSums(is.na(featuresDataframe)) == 0,]

```